### 2020-03-25 probabilisticProgrammingMartingales.tex
Start writing about the construction of ranking functions
Loosen the definition of ranking function slightly so as to be more
convenient. Alter Thm 1 to match.

\section{Ranking functions}
Given a probabilistic program (i.e.~a term $M$), in order to construct a supermartingale to prove its a.s.~termination, a function to assign values to each reachable program state is necessary. Define a ranking function on $M$ to be a measurable function $f:Rch(M) \to \mathbb{R}$ such that
\begin{itemize}
    \item $f(N) > 0$ for all $N$ not in normal form
    \item $f(N) = 0$ for all $N$ in normal form
    \item $f(E[\tY N]) \geq 1 + f(E[N(\tY N)])$
    \item $f(E[\tsample]) \geq \int_0^1 f(E[\underline{x}]) \, \mathrm{d}x$
    \item $f(E[R]) \geq f(E[N])$ for any other redex $R$, where $R \to N$
\end{itemize}

@@

\begin{theorem}
  If an SPCF term has a ranking function, it terminates almost surely.
\end{theorem}
\begin{proof}
Given a term $M$ and a ranking function $f$ for it, define random variables on the probability space $S$ (where $s$ is a random variable) by
\begin{align*}
(M_n,s_n) & = \red^n(M,s) \\
y_0 & = 0 \\
y_{n+1} & = \min \{ k \mid k>y_n, M_k \text{ normal form or of the form } E[\tY N] \}\\
M'_n & = M_{y_n} \\
X_n & = f(M'_n)
\end{align*}
and define a filtration $\mathcal{F}_n = \sigma(M_k, k \leq n)$ (i.e. all the samples used up to step $n$).

\lo{So all the random variables just defined are constant functions. E.g.~$M_n$ is a constant function from $S$ to the measurable space of PPCF terms. \akr{I think we discussed this and it was resolved.}}

@@

Given $M'_n$, there is some finite bound on the number of reduction steps that can take place from $M'_n$ without a $\tY$-reduction step, because of the type system, therefore $y_{n+1}$ is (conditional on $\mathcal{F}_{y_n+1}$) a bounded stopping time, therefore $\mathbb{E}[f(M_{y_{n+1}}) \mid \mathcal{F}_{y_n+1}] \leq f(M_{y_n+1})$. If $f(M_{y_n}) > 0$, then $M_{y_n} = E[\tY N]$ for some $E, N$, therefore $M_{y_n+1} = N (\tY N)$ and $f(M_{y_n+1}) \leq f(M_{y_n}) - 1$, therefore if $X_n > 0$, $\mathbb{E}[X_{n+1} \mid \mathcal{F}_{y_n+1}] \leq X_n - 1$.

Overall, this implies that $\mathbb{E}[X_{n+1}] - \mathbb{E}[X_{n}] \leq -\mathbb P[X_n > 0]$ therefore as $\mathbb{E}[X_n]$ is bounded below, $\mathbb P[X_n > 0]$ must tend to 0 as $n \to \infty$. $X_n = 0 \Rightarrow X_{n+1} = 0$ therefore $\mathbb P[X_n > 0 \text{ for infinitely many values of n}] = \mathbb P[\forall n: X_n > 0] \leq \inf_n \mathbb P[X_n > 0] = 0$, therefore $M$ terminates almost surely.
\end{proof}

>>>>

\section{Ranking functions}
Given a probabilistic program (i.e.~a term $M$), in order to construct a supermartingale to prove its a.s.~termination, a function to assign values to each reachable program state is necessary. Define a ranking function on $M$ to be a measurable function $f:Rch(M) \to \mathbb{R}$ such that
\begin{itemize}
    \item $f(N) \geq 0$ for all $N$
    \item $f(E[\tY N]) \geq 1 + f(E[N(\tY N)])$
    \item $f(E[\tsample]) \geq \int_0^1 f(E[\underline{x}]) \, \mathrm{d}x$
    \item $f(E[R]) \geq f(E[N])$ for any other redex $R$, where $R \to N$.
\end{itemize}

Any term for which a ranking function exists is called ``rankable''.

@@

\begin{theorem} \label{rankable implies ast}
  If an SPCF term is rankable, it terminates almost surely.
\end{theorem}
\begin{proof}
Given a term $M$ and a ranking function $f$ for it, define random variables on the probability space $S$ (where $s$ is a random variable) by
\begin{align*}
(M_n,s_n) & = \red^n(M,s) \\
y_0 & = 0 \\
y_{n+1} & = \min \{ k \mid k>y_n, M_k \text{ normal form or of the form } E[\tY N] \}\\
M'_n & = M_{y_n} \\
X_n & = f(M'_n)
\end{align*}
and define a filtration $\mathcal{F}_n = \sigma(M_k, k \leq n)$ (i.e. all the samples used up to step $n$).

@@

Given $M'_n$, there is some finite bound on the number of reduction steps that can take place from $M'_n$ without a $\tY$-reduction step, because of the type system, therefore $y_{n+1}$ is (conditional on $\mathcal{F}_{y_n+1}$) a bounded stopping time, therefore $\mathbb{E}[f(M_{y_{n+1}}) \mid \mathcal{F}_{y_n+1}] \leq f(M_{y_n+1})$. If $M_{y_n}$ isn't already in normal form, then $M_{y_n} = E[\tY N]$ for some $E, N$, therefore $M_{y_n+1} = N (\tY N)$ and $f(M_{y_n+1}) \leq f(M_{y_n}) - 1$, therefore if $M_{y_n}$ isn't normal form, $\mathbb{E}[X_{n+1} \mid \mathcal{F}_{y_n+1}] \leq X_n - 1$.

Overall, this implies that $\mathbb{E}[X_{n+1}] - \mathbb{E}[X_{n}] \leq -\mathbb P[M_{y_n} \text{not n.f.}]$ therefore as $\mathbb{E}[X_n]$ is bounded below, $\mathbb P[M_{y_n} \text{not n.f.}]$ must tend to 0 as $n \to \infty$. $M_{y_n} \text{n.f.} \Rightarrow M_{y_{n+1}} \text{n.f.}$ therefore $\mathbb P[M_{y_n} \text{not n.f.~for infinitely many values of n}] = \mathbb P[\forall n: M_{y_n} \text{not n.f.}] \leq \inf_n \mathbb P[M_{y_n} \text{not n.f.}] = 0$, therefore $M$ terminates almost surely.
\end{proof}

\section{Constructing Ranking Functions}
Although rankability implies almost sure termination, the converse does not hold in general. For example,
\begin{equation}
\tif{-\tsample < 0}{\underline{0}}{\tY \lambda x. x}
\end{equation}
terminates in 3 steps with probability 1, but isn't rankable because $\tY \lambda x. x$ is reachable, although that has probability 0. Not only is this counterexample a.s.t, it's positively almost surely terminating i.e.~the expected time to termination is finite.

A ranking function can be constructed under the stronger assumptions that, for every $N$ reachable from $M$, the expected number of $\tY$-reduction steps from $N$ to a normal form is finite. In particular, the expected number of $\tY$-reduction steps from each reachable term is a ranking function.

\akr{I don't know whether you'll see this before I actually write the other bit, but I'm planning on giving a more practical way of constructing ranking functions later using a similar technique, which is why this theorem seems relevant.}
\begin{theorem} \label{minimal}
Given a term $M$, the function $f:Rch(M) \to \mathbb R$ given by $f(N) = \mathbb E [\text{the number of }\tY\text{-reduction steps from }N\text{ to a normal form}]$, if it exists, is the least of all possible ranking functions of $M$.
\end{theorem}
\begin{proof}
Let $f$ be the candidate least ranking function defined above, and suppose $g$ is another ranking function such that $f(N) > g(N)$ for some $N \in Rch(M)$. The difference $g - f$ is then a supermartingale (with the same setup as in the proof of Theorem \ref{rankable implies ast}. The restrictions of $f$ and $g$ to $Rch(N)$ have the same properties assumed of $f$ and $g$, so assume w.l.o.g.~that $N=M$.

\akr{TODO: finish this proof. It shouldn't be complicated.}
\end{proof}

\paragraph{}
Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and Theorem \ref{minimal} makes even stronger assumptions than almost sure termination, so it isn't useful for proving it. Take, for example, the term
\begin{align*}
&(\tY \lambda f, n: \\
&\quad \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
&) \underline{0}
\end{align*}
which generates a geometric distribution.
    Despite its simplicity, its $Rch$ contains all the terms
\begin{itemize}
    \item $(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) \underline{i}$
    \item $(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: tif{\tsample - 0.5 < 0}{n}{f (n+1)}) \underline{i}$
    \item $(\lambda n: \tif{\tsample - 0.5 < 0}{n}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{n}{f (m+1)}) (m+1)}) \underline{i}$
    \item $\tif{\tsample - 0.5 < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
    \item $\tif{\underline r - 0.5 < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
    \item $\tif{\underline{r-0.5} < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
    \item $\underline{i}$
    \item $(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)$.
\end{itemize}
\akr{TODO: make this into a diagram with arrows.}
Even in this simple case, defining a ranking function explicitly is awkward because of the number of cases, although in most cases, because the value need only be greater than or equal to that of the next term in sequence, it suffices to take the ranking function as taking only 3 distinct values.
